Compositional Verification of Concurrent C Programs with Search Structure Templates

### 2023-04-27 concurrency/CONCUR 2023/paper.tex
clarified the interface mismatch around inRange

We will discuss several templates in section~\ref{templates}, but broadly, a search structure template consists of a \lstinline{traverse} function that travels through a data structure in search of a particular key, using the appropriate concurrency control mechanism when moving from node to node. The node to travel to is selected by a black-box function \lstinline{findNext} provided by the data structure; all the template needs to know is that it has some way of choosing a next node to examine when the current node does not contain the target key. \wm{Show an example template here?} Once the appropriate node for the key has been found, the template returns it to a top-level function that calls out to the data structure to perform the appropriate operation (e.g. insert, lookup, delete) on that node. Thus, the \lstinline{traverse} function can be verified without knowing anything about the target data structure other than its synchronization mechanism, and without knowing anything about the operation to be performed other than that it can be performed on a single node (or sometimes a node and its parent). The job of the data structure is to implement the \lstinline{findNext} function and the specific operations; the combination of the template and the data structure is then a fully operational concurrent data structure. Krishna et al.~presented templates in terms of \emph{flow interfaces}~\cite{krishna2017flow}, a framework for reasoning about graph-structured data structures, but many common search structures are not graphs; in this paper, we present a version of templates that is independent of flow interfaces. %is this a contribution in itself?
%what more do we want to say here?

@@

To apply the template approach to real-world code, we instead use the Verified Software Toolchain (VST)~\cite{plcc}, a separation-logic-based verifier for C programs. VST is also built in Coq and is connected to the CompCert verified C compiler~\cite{compcert}, giving it a realistic semantics and a guarantee that proved properties will hold on compiled code. Recent work on VST~\cite{vst-iris} extended it to support most of the advanced concurrency features of Iris, including ghost state, invariants, and atomic specifications. Building on this, we were able to reconstruct the template approach in VST and apply it to real C programs. %clean this up a little

@@

\subsection{Search Structure Templates}
\label{templates}
The C implementation of a node in our template style is:
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
typedef struct node_t {node *t; lock_t *lock;} node_t;
\end{lstlisting}

Each node of type \lstinline{node_t} has a \lstinline{lock} field that protects the node and a \lstinline{t} field that points to a \lstinline{node} struct. Depending on the data structure, the \lstinline{t} field can have different interpretations. In a binary search tree, if the \lstinline{t} field is \lstinline{NULL}, it indicates a leaf node with no key or value; otherwise, it points to a \lstinline{node} struct containing the node's key, value, and child pointers. 
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
typedef struct node {int key; void *value; struct node_t *left, *right;} node;
\end{lstlisting}

For a linked list, a \lstinline{t} field with \lstinline{NULL} represents the last node in the list, while a non-null \lstinline{t} field points to a \lstinline{node} struct holding the node's key, value, and a pointer to the next item in the list.

\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
typedef struct node {int key; void *value; struct node_t *next;} node;
\end{lstlisting}

%For the purpose of reasoning about lock-coupling style (also known as hand-over-hand locking), we specifically consider traversing the tree using this locking method. To accomplish this, we must maintain at least one lock during traversal between nodes. This requires obtaining the lock for the next node before releasing the lock for the current node. Therefore, we utilize the \lstinline{pn} struct containing two \lstinline{node_t} type nodes: the \lstinline{p} field representing the current node, and the \lstinline{n} field denoting the next node.

\subsubsection{Lock-coupling Template}
\label{lock-coupling-algo}
The lock-coupling template employs the hand-over-hand locking scheme to prevent interference from other threads during traversal. Each thread continuously maintains at least one lock during traversal between nodes, ensuring that other threads cannot interfere with the traversal or invalidate the ongoing search.

Figure \ref{traverse_lock} shows the \lstinline{traverse} function as presented by Krishna et al.~(Figure \ref{traverse_lock_a}) and the corresponding C implementation (Figure \ref{traverse_lock_b}). The C implementation uses a struct
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
typedef struct pn {struct node_t *p; struct node_t *n;} pn;
\end{lstlisting}
to mimic the pair of nodes \lstinline{(p, n)} returned by the functional implementation, where \lstinline{n} is the current node and \lstinline{p} is its parent. The algorithm assumes that certain helper functions must be provided by the implementation that satisfy certain criteria. One such function, namely \lstinline{findNext}, is utilized by the algorithm to determine the next node $\texttt{n'}$ to be visited based on the current node $\texttt{n}$ and the key $\texttt{k}$. In the functional version, \lstinline{findNext} returns an \lstinline{option node}; in C, it instead returns a boolean and, if a next node is found, modifies \lstinline{pn->n}.

To ensure thread safety, it is essential to acquire and release locks in a specific order during traversal. The lock-coupling scheme achieves this by requiring threads to acquire locks in increasing order of node addresses. The thread then releases the lock for the previous node before acquiring the lock for the next node in the traversal sequence. This guarantees that the locks are acquired and released in the same order across all threads and prevents deadlocks from occurring.

@@

To implement the template methodology for concurrent data structure implementations in C, we have implemented the \lstinline{traverse} function in C, as depicted in Figure \ref{traverse_lock_a}.  Our version of the \lstinline{traverse} function closely resembles the one written in an ML-like language, as it relies on a helper function called \lstinline{findNext}. The purpose of \lstinline{findNext} is to determine the next node to be visited during the traversal. In our implementation, the \lstinline{traverse} function stops when it reaches the base case, which occurs when the current node is \lstinline{NULL}. Alternatively, it terminates if \lstinline{findNext} fails to find a next node that includes the key value $\texttt{k}$ in the set of key values. The \lstinline{traverse} function returns a Boolean value indicating whether it has successfully found the next node during traversal. If it fails to do so, it returns false, which other functions can use to make decisions. For instance, in a binary search tree, a node with a value of 42 would have a left child with a set of key values consisting of values less than 42 and a right child with one greater than 42. If we want to insert a new node with the target key $\texttt{k}$ into the tree, we can use the \lstinline{traverse} function to traverse the tree. When \lstinline{traverse} fails to find the next node (either to the left or right), it implies that it has reached a node with a target key $\texttt{k}$ that equals the node's value.

When \lstinline{findNext} arrives at a new node, it acquires a lock for that node, as is the nature of hand-over-hand locking. It then releases the lock for the current node, ensuring that the link between the two nodes is not removed or rearranged while traversing it. To utilize the template methodology for concurrent data structure implementations, it is crucial to keep track of the locks acquired for each node during the traversal.

\subsubsection{Give-Up Template}
\label{give-up-algo}
The give-up template is an algorithm for implementing concurrent search in a tree data structure. Unlike the lock-coupling template, which maintains locks during traversal between nodes, the give-up template takes an \emph{optimistic concurrency control} approach \lb{c.f. RCU}, only acquiring locks when loading or storing to a node. This reduces the amount of time that locks are held and makes the algorithm more scalable in highly-contended scenarios. Each node has a range field that serves as a reference point for threads to determine if they are still on the correct path. If a thread finds that it has deviated from the expected path, it abandons the node by releasing the lock and returns to the root node to start again. This happens when another concurrent thread restructures the data structure, invalidating the state of the current thread in between releasing one lock and acquiring another. For instance, the node we are traveling to may be moved to another part of the data structure intended to hold a different set of keys, or even deleted, before we acquire its lock.

\begin{figure}[!ht]
	\begin{subfigure}[t]{0.45\textwidth}
		\lstinputlisting[language=caml, style=myStyle]{giveup_traverse.ml} 
		\caption{The \lstinline{traverse} method of the give-up template algorithm written in ML-like language}
		\label{traverse_giveup_a}	
	\end{subfigure}\qquad
	\begin{subfigure}[t]{0.48\textwidth}
		\lstinputlisting[language=C, style=myStyle]{giveup_traverse.c} 
		\caption{The \lstinline{traverse} method of the give-up template algorithm written in C}
		\label{traverse_giveup_b}
	\end{subfigure}
	\caption{The \lstinline{traverse} method of the give-up template algorithms}
	\label{traverse_giveup}
\end{figure}

Figure \ref{traverse_giveup_a} depicts the give-up template algorithm as originally presented. The give-up template algorithm uses a helper function called \lstinline{inRange}, in addition to the \lstinline{findNext} function discussed in the previous section. The purpose of the \lstinline{inRange} function is to determine whether the key value $\texttt{k}$ falls within the range of keys held in node $\texttt{n}$ and its successors. If the key value is outside the node's range (e.g. because the node has been relocated), the search is terminated, and the thread is restarted. To implement the template algorithm, it is crucial to keep track of the lower and upper bounds of the keys present in the subtree rooted at the current node. Therefore, we have incorporated two new fields, \lstinline{min} and \lstinline{max}, into each node of type \lstinline{node_t} in our C implementation. These fields store the lower and upper bounds of the keys, respectively. In the context of binary search trees, each node is labeled with a range that is associated with a lower and upper bound on the keys. This begins with $(\texttt{min}, \texttt{max}) = (-\infty, +\infty)$ at the root node and is propagated to the empty leaf nodes. If the root node (which also serves as the parent of a node) has a key of $42$, the left leaf node's range would be $(\texttt{min}, \texttt{max}) = (-\infty, 42)$, and the right leaf node's range would be $(\texttt{min}, \texttt{max}) = (42, +\infty)$.

%\wm{Describe why we need extra pieces in the struct, and how this relates to the other version.}
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
	typedef struct node_t {node *t; lock_t *lock; int min; int max;} node_t;
\end{lstlisting}

>>>>

We will discuss several templates in section~\ref{templates}, but broadly, a search structure template consists of a \lstinline{traverse} function that travels through a data structure in search of a particular key, using the appropriate concurrency control mechanism when moving from node to node. The node to travel to is selected by a black-box function \lstinline{findNext} provided by the data structure; all the template needs to know is that it has some way of choosing a next node to examine when the current node does not contain the target key. \wm{Show an example template here, and enough of the specification to explain atomicity} Once the appropriate node for the key has been found, the template returns it to a top-level function that calls out to the data structure to perform the appropriate operation (e.g. insert, lookup, delete) on that node. Thus, the \lstinline{traverse} function can be verified without knowing anything about the target data structure other than its synchronization mechanism, and without knowing anything about the operation to be performed other than that it can be performed on a single node (or sometimes a node and its parent). The job of the data structure is to implement the \lstinline{findNext} function and the specific operations; the combination of the template and the data structure is then a fully operational concurrent data structure. Krishna et al.~presented templates in terms of \emph{flow interfaces}~\cite{krishna2017flow}, a framework for reasoning about graph-structured data structures, but many common search structures are not graphs; in this paper, we present a version of templates that is independent of flow interfaces. %is this a contribution in itself?
%what more do we want to say here?

@@

To apply the template approach to real-world code, we instead use the Verified Software Toolchain (VST)~\cite{plcc}, a separation-logic-based verifier for C programs. VST is also built in Coq and is connected to the CompCert verified C compiler~\cite{compcert}, giving it a realistic semantics and a guarantee that proved properties will hold on compiled code. Recent work on VST~\cite{iris-vst-arxiv} extended it to support most of the advanced concurrency features of Iris, including ghost state, invariants, and atomic specifications. In this paper, we use these extensions to reconstruct the template approach in VST and apply it to real C programs. %clean this up a little

@@

\subsection{Defining the Templates}
\label{templates}

The core of each search structure template is a \lstinline{traverse} function that navigates a data structure using the target synchronization mechanism. Templates abstract away from the data structure being traversed, operating on a generic \lstinline{node} type and taking data-structure-specific functions like \lstinline{findNext} (which chooses the next node to search) as parameters. We consider lock-based synchronization in this paper, and so our generic node type can be defined in C as:
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
typedef struct node_t {node *t; lock_t *lock;} node_t;
\end{lstlisting}
There are two things worth noting about this data structure. First, each node has its own lock: all of our templates will use \emph{fine-grained locking}, holding as few locks as possible to allow other parts of the data structure to be modified concurrently. Second, the \lstinline{node} struct itself is defined by the target data structure: for instance, a binary search tree might define it as 
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
typedef struct node {int key; void *value; struct node_t *left, *right;} node;
\end{lstlisting}
while a linked list might use
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
typedef struct node {int key; void *value; struct node_t *next;} node;
\end{lstlisting}
This is the first place where we decompose our program into a generic concurrent component and a sequential data-structure component. Now, we use \lstinline{node_t} to define \lstinline{traverse} functions for specific concurrency patterns.

\subsubsection{Lock-Coupling Template}
\label{lock-coupling-algo}

The first template we consider is lock coupling (also called hand-over-hand locking), in which threads use the locks on each node to prevent interference from other threads during traversal. Each thread always holds at least one lock, and acquires the lock on the next node before releasing its current lock, ensuring that other threads cannot invalidate the ongoing search.
Figure \ref{traverse_lock} shows the lock-coupling \lstinline{traverse} function as presented by Krishna et al.~(Figure \ref{traverse_lock_a}) and our corresponding C implementation (Figure \ref{traverse_lock_b}). The C implementation uses a struct
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
typedef struct pn {struct node_t *p; struct node_t *n;} pn;
\end{lstlisting}
to mimic the pair of nodes \lstinline{(p, n)} returned by the functional implementation, where \lstinline{n} is the current node and \lstinline{p} is its parent. The template relies on one function provided by the underlying data structure, namely \lstinline{findNext}, which is used to determine the next node $\texttt{n'}$ to be visited based on the current node $\texttt{n}$ and the key $\texttt{k}$. In the functional version, \lstinline{findNext} returns an \lstinline{option node}; in C, it instead returns a boolean and, if a next node is found, modifies \lstinline{pn->n}.

%To ensure thread safety, it is essential to acquire and release locks in a specific order during traversal. The lock-coupling scheme achieves this by requiring threads to acquire locks in increasing order of node addresses. The thread then releases the lock for the previous node before acquiring the lock for the next node in the traversal sequence. This guarantees that the locks are acquired and released in the same order across all threads and prevents deadlocks from occurring.

@@

Our implementation of \lstinline{traverse} translates the functional implementation into more idiomatic C code. The lock-coupling pattern can be seen on lines 14 and 15, where \lstinline{traverse} acquires the next node's lock and then releases the current node's lock. The function stops when it reaches an empty leaf node (\lstinline{pn->p->t = NULL}), or when \lstinline{findNext} returns 0, indicating that the target key \lstinline{k} is in the current node. The \lstinline{traverse} function returns an integer flag that is 1 when we reach a leaf node and 0 when we find \lstinline{k} in an existing node; the behavior of operations that call \lstinline{traverse} can vary depending on this flag. For instance, an \lstinline{insert} operation may create a new node when the flag is 1 and modify an existing node when the flag is 0, while a \lstinline{lookup} operation may fail when the flag is 1 and return the value in the current node when the flag is 0.

%When \lstinline{traverse} arrives at a new node, it acquires the lock for that node and then releases the lock for the current node, according to the lock-coupling pattern---this ensures that the link between the two nodes is not removed or rearranged while traversing it. To utilize the template methodology for concurrent data structure implementations, it is crucial to keep track of the locks acquired for each node during the traversal.

\subsubsection{Give-Up Template}
\label{give-up-algo}

We next consider the give-up template, which uses an \emph{optimistic concurrency control} approach, acquiring fewer locks at the cost of sometimes having to recover from synchronization errors. Unlike the lock-coupling template, which maintains locks during traversal between nodes, the give-up template only acquires a lock just before operating on a node, and holds at most one lock at any time. This means that a conflicting operation may invalidate our traversal, for instance by moving the next node to another part of the data structure before we acquire its lock. To guard against this, the give-up template maintains a \emph{range} field on each node indicating the range of keys that should be stored in this node and its successors, and \lstinline{traverse} must check the range field at each node to ensure that it is still on the right path. If a check fails, we give up and start the traversal over from the root node. The give-up template performs well in scenarios where operations generally do not conflict, either because they are mostly to independent parts of the data structure or because they mostly do not delete or relocate nodes.

\begin{figure}[!ht]
	\begin{subfigure}[t]{0.45\textwidth}
		\lstinputlisting[language=caml, style=myStyle]{giveup_traverse.ml} 
		\caption{The \lstinline{traverse} method of the give-up template algorithm written in an ML-like language}
		\label{traverse_giveup_a}	
	\end{subfigure}\qquad
	\begin{subfigure}[t]{0.48\textwidth}
		\lstinputlisting[language=C, style=myStyle]{giveup_traverse.c} 
		\caption{The \lstinline{traverse} method of the give-up template algorithm written in C}
		\label{traverse_giveup_b}
	\end{subfigure}
	\caption{The \lstinline{traverse} method of the give-up template algorithms}
	\label{traverse_giveup}
\end{figure}

Figure \ref{traverse_giveup_a} depicts the give-up template algorithm as originally presented. In addition to \lstinline{findNext}, the \lstinline{traverse} function uses a helper function called \lstinline{inRange}, which determines whether the key value $\texttt{k}$ falls within the range of keys held in node $\texttt{n}$ and its successors. If \lstinline{k} is outside the node's range (e.g. because the node has been relocated), the search is restarted from the root node \lstinline{r}.

The \lstinline{inRange} function raises an interesting question about the template approach: is \lstinline{inRange} part of the give-up template, or the underlying data structure? Some data structures may already track the range of keys expected in the current node, and so might define \lstinline{inRange} even in sequential implementations. However, in most sequential settings the key ranges can be computed from the data structure (e.g., in a binary search tree, the left subtree of a node with key \lstinline{k} holds keys less than \lstinline{k}), and there is no reason to explicitly store a node's range in the node itself. The give-up template of Krishna et al.~implicitly assumes that the underlying data structure supports \lstinline{inRange}; we prefer to consider \lstinline{inRange} part of the template, so that the template can be applied to as many target data structures as possible without modifying them. Accordingly, in this template we add two new fields, \lstinline{min} and \lstinline{max}, to the type \lstinline{node_t} in our C implementation:
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
	typedef struct node_t {node *t; lock_t *lock; int min; int max;} node_t;
\end{lstlisting}
These fields store lower and upper bounds on the keys reachable from the current node. %In the context of binary search trees, each node is labeled with a range that is associated with a lower and upper bound on the keys. This begins with $(\texttt{min}, \texttt{max}) = (-\infty, +\infty)$ at the root node and is propagated to the empty leaf nodes. If the root node (which also serves as the parent of a node) has a key of $42$, the left leaf node's range would be $(\texttt{min}, \texttt{max}) = (-\infty, 42)$, and the right leaf node's range would be $(\texttt{min}, \texttt{max}) = (42, +\infty)$.
We can then define \lstinline{inRange} as a helper function in the template, rather than requiring data structures to provide it. \wm{This is the first but not the last implicit assumption in the templates; can we say something more general?}
